\LARGE \bf Effective Nonparametric Distribution Modeling \\ for Distribution Approximation Applications

### 2020-02-13 [2019-02]_NUMA_(interpolant_error_bound)/response_1.txt

Response to Reviewers

Thank you all for the thoughtful reviews and comments on the paper. We
have revised the paper and believe the revisions address all concerns.
The major endowed improvements include:

  (1) a new proof for lemma 3 that results in a sharp bound on the
      error of the gradient approximation (Section 5, page 10),

  (2) a new example that demonstrates the sharpness of the theoretical
      error bound (Section 5, page 11),

  (3) a new subsection illustrating the modeling accuracy of a
      regression technique and an interpolation technique on a known
      analytic test function when given an increasing number of
      well spaced approximation points (Section 5.1, pages 12 and 13),

  (4) a radial basis function kernel support vector regressor (SVR)
      was substituted in place of the polynomial kernel SVR,

  (5) a new minor Section in the appendix defines relevant statistical
      terminology used throughout the paper (Appendix, page 25).

Responses to individual requests / comments by reviewers follows.

  * missing regression based on radial basis functions

Addressed by (4), updated description in last paragraph of Section 2.3
(page 4) to coincide with (4).

  * missing regression / interpolation based on sparse grids

The box spline method described in Section 3.4 uses a sparse grid to
construct its approximation (described in detail in reference [33]).
The phrase "sparse grid" was added to the first sentence of 3.4 to
make this more clear.

  * missing interpolation over [approximate Fekete points]

Addressed by (3), the "well spaced" points used for model construction
are approximate Fekete points. See Section 5.1 paragraph 3, page 12.

  * missing numerical evidence about the error bound from the theorem
    in Section 5, is it sharp?

Addressed by (2), Section 5 page 11 paragraph 2. An example shows that
(with modified lemma 3, per (1)), the error bound is sharp.

  * would expect some convergence plot showing error decreasing with
    more points

Included in (3), Section 5.1 Figures 2 and 3, pages 12 and 13.

  * missing a comparison on simulated data
  * recommend tests approximating functions with analytic expression
    (refer to Barthelmann et al. 2000)

Included in (3), Section 5.1 pages 12 through 14. Used a modified
version of the "Oscillatory" test function from provided reference.
The original dot product structure of the test function strongly
resembles the internal structure of the multilayer perceptron. The
slight modification was made to reduce bias in the comparison.

  * should include comparison on simulated data both with and without
    randomness incorporated into the data

Addressed by (3), where the left and right halves of Figures 2 and 3
in Section 5.1 (pages 12 and 13) now depict the effect randomness has
on approximation error.

  * in case the models are not approximating CDFs, alternatives to the
    KS statistic should be discussed

A sentence was added to the last paragraph of Section 7 that mentions
the quantification of error with alternative measures.

  * right now the box-plots visualize the KS statistic, not the KS
    measure

The KS statistic is the max norm difference (the KS measure). This was
clarified in the caption of Figure 13 (page 19) and the second
paragraph of Section 6.5 (page 19). We refer to the visual in Figure 1
for further clarification, which illustrates that the KS statistic is
the max norm difference between two cumulative distribution functions.

  * readers may have a weaker background in statistics, terms like
    null hypothesis, p-value, and box-plots cannot be given for
    granted

Addressed by (5), which is also now referred to in the last sentence
of Section 4 (page 8).

  * statistical discussion in the Figures should be moved to main text

This suggestion is appreciated. Revision (5) was preferred for fear
that removing important definitions from Figure text may reduce
comprehension for less thorough readers of the paper.

  * state clearer in the introduction which are the main objectives
    and contributions of this work

The second to last paragraph of Section 1 (page 3) has been modified
to clearly state both the objectives and contributions of this work.

### 2020-02-13 [2019-12]_IEEESE_(quintic_applied)/paper.tex
Updated IEEESE paper with revisions suggested by Dr Hong. Added backwards compatible 'util' subdirectory to the NUMA paper to begin working on revisions.

%% \author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2,5}$, Tyler H. Chang$^{1}$, Li Xu$^{3}$, Yueyao Wang$^{3}$,\\ Jon Bernard$^{1}$, Yili Hong$^{4}$, Kirk W. Cameron$^{2}$% <-this % stops a space
%% \thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
%% \thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
%% \thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% \thanks{$^{3}$Doctoral candidate, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% \thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% \thanks{$^{5}$Departments of Mathematics and Aerospace Engineering, Virginia Polytechnic Institute and State University, Virginia 24060, USA}%
%% }

@@

Many fields of science rely on the collection of samples and estimation of true population distributions from those samples. There are several effective nonparametric methods for approximating a true distribution from empirical data, however it is unclear which methods produce the best approximations in practice. This work presents a case study on the effectiveness of various distribution approximations. Results show that piecewise linear approximations produce the smallest maximum absolute error, while the classic empirical distribution function (EDF) produces the smallest median absolute error as well as the smallest absolute error at lower percentiles when approximating a distribution from a sample. When building distribution prediction models, the piecewise quintic and cubic approximations produce the lowest absolute error at most error percentiles. These initial results encourage more research on the best methods of fitting empirical data with smooth functions to generate accurate distribution approximations.

@@

In this statistical work, the construction of a monotone interpolating spline that is continuous in its derivatives could be meaningfully useful \cite{ramsay1988monotone}. A function with $C^1$, and especially $C^2$ continuity could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals. A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a desirable property many standard parametric distributions maintain.

There is publicly available software for monotone piecewise polynomial interpolation, including quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}. Recently an algorithm for the construction of monotone quintic splines has been produced as well (citation pending).%% \cite{lux2020algorithm}
This work considers the EDF, a linear interpolant, a monotone cubic spline interpolant, and a monotone quintic spline interpolant as candidate approximations.

@@

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

@@

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.5\textwidth]{delaunay-predictions.pdf}
  %% \includegraphics[width=.5\textwidth]{abs-errors-100-samples.pdf}
  \caption{The result of using Delaunay to predict distributions. Notice that the best performing, lowest error (to the left), approximation is the EDF. Interestingly, the piecewise linear fit does not produce results as accurate as the EDF. This likely has to do with the inherent smoothing that happens when multiple different EDFs are combined to make one prediction.
  \vspace{-.5cm}}
  \label{fig:delaunay-errors}
\end{figure}

>>>>

\author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2,5}$, Tyler H. Chang$^{1}$, Li Xu$^{3}$, Yueyao Wang$^{3}$,\\ Jon Bernard$^{1}$, Yili Hong$^{4}$, Kirk W. Cameron$^{2}$% <-this % stops a space
\thanks{*This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.}%
\thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
\thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{3}$Doctoral candidate, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{5}$Departments of Mathematics and Aerospace Engineering, Virginia Polytechnic Institute and State University, Virginia 24060, USA}%
}

@@

Many fields of science rely on the collection of samples and estimation of true population distributions from those samples. There are several effective nonparametric methods for approximating a true distribution from empirical data, however it is unclear which methods produce the best approximations in practice. This work presents a case study on the effectiveness of various distribution approximations. Results show that piecewise linear approximations produce the smallest maximum absolute error, while the classic empirical distribution function (EDF) produces the smallest median absolute error as well as the smallest first quartile and minimum absolute error when approximating a distribution from a sample. When building distribution prediction models, the piecewise quintic and cubic approximations produce the lowest absolute error at most error percentiles. This case study encourages more research on the best methods of fitting empirical data with smooth functions to generate accurate distribution approximations.

@@

In this statistical work, the construction of a monotone interpolating spline that is continuous in its derivatives could be meaningfully useful \cite{ramsay1988monotone}. A function with $C^1$, and especially $C^2$ continuity could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals. A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a desirable property many standard parametric distributions maintain. In effect, an improved distribution estimate can increase the accuracy of distribution predicting models \cite{xu2020modeling}.

There is publicly available software for monotone piecewise polynomial interpolation, including quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}. Recently an algorithm for the construction of monotone quintic splines has been produced as well \cite{lux2020algorithm}.
This work considers the EDF, a linear interpolant, a monotone cubic spline interpolant, and a monotone quintic spline interpolant as candidate approximations.

@@

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{example_distribution_prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.1, .3, .6\} (top to bottom at middle). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (vertical line) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

@@

\begin{figure}
  \vspace{-.3cm}
  \includegraphics[width=.5\textwidth]{delaunay-predictions.pdf}
  %% \includegraphics[width=.5\textwidth]{abs-errors-100-samples.pdf}
  \caption{The result of using Delaunay to predict distributions. Notice that the best performing, lowest error (to the left), approximations at almost all absolute error percentiles are the cubic and quintic interpolants, closely followed by the linear interpolant. Interestingly, these results are contrary to pure distribution approximations from a sample. This likely has to do with the inherent smoothing that happens when multiple different approximations are combined to make one prediction.
  \vspace{-.5cm}}
  \label{fig:delaunay-errors}
\end{figure}
